Theoretical prediction of disrupted Min oscillation in flattened \emph{Escherichia coli}

### 2015-04-01 paper/plos.tex
fix Mannik spelling Tested everything on bennet.

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf\newline{Theoretical prediction of disrupted Min oscillation in
  flattened \emph{Escherichia coli}}
}
\newline
% Insert Author names, affiliations and corresponding author email.
\\
Jeff B. Schulte\textsuperscript{1,*},
Rene W. Zeto\textsuperscript{1},
David Roundy\textsuperscript{1}
\\
\bf{1} Dept. of Physics/Oregon State University/Corvallis, Oregon, USA
\\

@@

Previous experimental studies have shown that the MinD protein system
is capable of exhibiting oscillations in round
shapes~\cite{corbin2002exploring,fange2006noise} as well as in
connected three pronged tube shapes~\cite{varma2008min}, in which the
oscillations seem to seek out the extreme poles in the cell.
%%   One conclusion of these
%% studies has been that the MinD system robustly forms bipolar
%% oscillations regardless of variations in cellular shape.
Mannik \emph{et al.} have recently shown that there are limitations to
this robust capability to
oscillate~\cite{mannik2012robustness,mannik2009bacterial,mannik2010bacteria}. They
have experimentally forced \emph{E. coli} cells into microfabricated
channels of thickness less than that of the natural diameter of the
cells, in which they area able to grow and divide. Upon entering the
channels, the cells undergo a mechanical deformation in which they
both widen and lengthen within the plane of the channel.  This
deformation results in very wide (they can reach widths of over
$5\micron$) flattened cells that when viewed from the top down have
irregular and asymmetric shapes.  While these cells are still able to
divide into surprisingly equal volumes, the MinD oscillations in these
cells are spatially irregular. Seen with fluorescent microscopy, the
MinD maximize in multiple locations within the cell, in a seemingly
random sequence. These experiments allow for an opportunity to test
models of the MinD system against more extreme cases than have been
seen thus far.  An additional interesting question is whether the
irregularity observed by Mannik \emph{et al.}  indicates that the MinD
system can become chaotic (as in
e.g. Ref.~\citenum{kuramoto1978diffusion}), or if this irregularity is
the result of a stochastic but nonchaotic process.
%

@@

%% biology~\cite{lawson2013spatial, robb2014stochastic,
%%   oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
In this paper we use Huang's original 2003
model~\cite{huang2003dynamic}, which has been widely
used~\cite{hattne2005stochastic,huang2004min,kerr2006division} in both
deterministic and stochastic variants, to study flattened cells of
$0.4\micron$ thickness that are similar to those observed by Mannik
\emph{et al.}~\cite{mannik2009bacterial}.

@@

We mean to investigate the geometric limits of the Min system
oscillations as observed by Mannik \emph{et
  al.}~\cite{mannik2012robustness}, so we have modeled the Min system
in several cell shapes and sizes.  Here we present a selection of
these, beginning with naturally occurring pill-shaped cells, followed
by a number of flattened out shapes which reflect the experiments of
Mannik \emph{et al.}, in which bacteria are confined within thin
slits.  These slits are fabricated with a width of .25\micron, but
they are coated with a PDMS lining that the cells are able to deform,
raising uncertainty in the actual cell thickness.  Tests with pure
silicon, non-deformable slits show that the smallest thickness for
which cells are able to penetrate is
.4\micron~\cite{mannik2012robustness,mannik2009bacterial}.  We
therefore assume that the PDMS slits have been deformed to this
thickness and simulate flattened cells with a thickness of .4\micron.
Viewed from the top down the cells have the shapes described below and
viewed from the side they have at their edges a semicircular
cross-section (one may imagine the shape of a pancake).
%
In this paper we focus on four specific flattened cell shapes.  Two of
these shapes replicate those published by Mannik \emph{et al.}, and
the other two are `stadium' shapes that respectively have the same
aspect ratio, thickness, and volume as the two cell shapes
experimentally observed by Mannik \emph{et
  al.}~\cite{mannik2012robustness}.  Viewed from the top down, these
stadium shapes appear as rectangles with semi-circular end caps on the
long axis ends.

@@

Figure~\ref{image-p} shows a series of color plots of the density of
proteins at each stage of the reaction cycle. Deterministic simulation
data is shown above and stochastic simulation data is shown below.
The cells shown are $4\micron$ in length, measured from end to end.
We have `smeared out' the stochastic concentrations in a manner meant
to reproduce the images shown by diffraction limited fluorescence
microscopy.  We do so using the two dimensional Gaussian approximation
developed by Zhang \emph{et al.}~\cite{zhang2007gaussian}.  In this
approximation we use a numerical aperture value of 1.3, which is the
same as used by Mannik experimentally, and a wavelength of $650nm$.
Each frame is 2.5 seconds ahead of the last, and each image shows the
concentration of a given state of protein (of the five described in
the reaction model) summed over the coordinate normal to the page.

@@

\begin{figure*}
  \centering
  \hspace{-3cm}
  \vspace{1cm}
  \includegraphics[width=16cm]{../paper/plot-ave}
  \caption{We display here arrows depicting successive maxima in space
    and and time overlayed on a color plot of the total MinD density
    averaged over the same time period.  The simulation time covered
    for the flattened cells is 350 seconds, which is the same
    period of time depicted in the experimental data plots of Mannik
    \emph{et al.}~\cite{mannik2012robustness}.  For the wild-type pill
    shape, we only cover 250 seconds, in order to provide a
    useful comparison due to its shorter oscillation period.  The top
    row shows plots published by Mannik \emph{et al.} of the MinD maxima behavior
    and the bottom two rows show our simulations using the stochastic
    and deterministic models, respectively.  We simulated
    approximations to the two shapes observed by Mannik, which we call
    \emph{shape A} and \emph{shape B}.  In addition, we studied two
    flattened stadium shapes which we call \emph{stadium A} and
    \emph{stadium B} corresponding in aspect ratio and thickness to
    the two experimental shapes.  The spatial length scale of all the
    figures shown was identical. Each of the flattened cell shapes
    uses the same color scale for the number of proteins per unit
    area.  Finally, we display the natural pill shape, which was also
    featured in Figs.~\ref{image-p} and~\ref{corr-pill}, with a
    different color scale to reflect the thicker cell containing more
    proteins per cross-sectional area.  }
  \label{randst-plot-ave}
\end{figure*}

@@

\subsection*{Comparison of Experimental and Stadium Shapes}
%\label{sec:model-method-shapes}
As explained above, we focus on simulation of four illustrative
flattened cell shapes: two shapes created to replicate those shapes
observed experimentally by Mannik \emph{et
  al.}~\cite{mannik2012robustness} (shape~\emph{A} and
shape~\emph{B}), and for comparison two `stadium' shapes that have the
identical aspect ratio, thickness and volume (stadium \emph{A} and
stadium \emph{B}), shown in Figure~\ref{randst-plot-ave}.  All of the
flattened cells have lost the rotational symmetry of the original pill
shapes, but while shape~\emph{A} and shape~\emph{B} have only one
mirror plane symmetry (that is normal to the flattened plane),
stadium~\emph{A} and stadium~\emph{B} have three.  This allows us to
distinguish between the effect of flattening the cell and the cell
shape's irregularity and asymmetry.

@@

From each of these flattened simulations, we have chosen a typical 350
second segment to compare with the published results of
Mannik \emph{et al.}~\cite{mannik2012robustness}, exemplified in arrow plots in
which the arrow heads show the location of sequential MinD maxima
within the cell (in Fig.~\ref{randst-plot-ave}).  In addition to
arrows between successive maxima in space and time, we plot as a
colored background the density of MinD proteins averaged over the same
time period.  We note that we have manually verified that our
(computer-generated) arrow plots also reflect a human interpretation
of a movie of the same data.  Finally, for comparison we present the
same plot for a wild-type cell, with a time period of 250 seconds to
account for its short period of oscillation.

@@

%% We found that the deterministic simulations of the MinD system results
%% in robust and regular oscillatory behavoir of polar selection and
%% oscillation in not only symmetrical shapes, such as the wildtype pill
%% shape, but also in very assymetrical, flatttened cells, such as those
%% observed by Mannik.  We also examined larger and smaller versions of
%% these shapes, and found this behavior to be very robust, from the
%% minimum size to sustain oscillation (around \fixme{???} $\mu$m) up to
%% around twice the cell size reported by Mannik.  At larger sizes than
%% this, less regular oscillations occur using the deterministic method,
%% but we note that the cells reported by Mannik are already considerably
%% larger in volume than typical wild type
%% cells~\cite{kubitschek1990cell,kubitschek1968linear,mannik2012robustness}.

@@

We find that flattened but regular and symmetric cells exhibit MinD
oscillations that are qualitatively similar to the oscillations
observed experimentally (and in our simulation) in irregular and
asymmetric cells.  This demonstrates that asymmetry is not required in
order to induce spatially irregular MinD oscillations.  In addition,
the temporal regularity of oscillations is only moderately affected by
irregularity in the shape of flattened cells.  Thus we conclude that
it is the flattening of the cells, and the accompanying expansion in the
transverse direction, rather than their irregular shape that primarily
causes the disruption of regular bipolar oscillation which is observed
by Mannik \emph{et al.}~\cite{mannik2012robustness}.

>>>>

% Title must be 150 characters or less
\begin{flushleft}
{\Large
\textbf\newline{Theoretical prediction of disrupted Min oscillation in
  flattened \emph{Escherichia coli}}
% Disrupted Min oscillation in flattened E. coli
}
\newline
% Insert Author names, affiliations and corresponding author email.
\\
Jeff B. Schulte\textsuperscript{1,*},
Rene W. Zeto\textsuperscript{1},
David Roundy\textsuperscript{1}
\\
\bf{1} Dept. of Physics/Oregon State University/Corvallis, Oregon, USA
\\

@@

Previous experimental studies have shown that the MinD protein system
is capable of exhibiting oscillations in round
shapes~\cite{corbin2002exploring,fange2006noise} as well as in
connected three pronged tube shapes~\cite{varma2008min}, in which the
oscillations seem to seek out the extreme poles in the cell.
%%   One conclusion of these
%% studies has been that the MinD system robustly forms bipolar
%% oscillations regardless of variations in cellular shape.
M\"annik \emph{et al.} have recently shown that there are limitations to
this robust capability to
oscillate~\cite{mannik2012robustness,mannik2009bacterial,mannik2010bacteria}. They
have experimentally forced \emph{E. coli} cells into microfabricated
channels of thickness less than that of the natural diameter of the
cells, in which they area able to grow and divide. Upon entering the
channels, the cells undergo a mechanical deformation in which they
both widen and lengthen within the plane of the channel.  This
deformation results in very wide (they can reach widths of over
$5\micron$) flattened cells that when viewed from the top down have
irregular and asymmetric shapes.  While these cells are still able to
divide into surprisingly equal volumes, the MinD oscillations in these
cells are spatially irregular. Seen with fluorescent microscopy, the
MinD maximize in multiple locations within the cell, in a seemingly
random sequence. These experiments allow for an opportunity to test
models of the MinD system against more extreme cases than have been
seen thus far.  An additional interesting question is whether the
irregularity observed by M\"annik \emph{et al.}  indicates that the MinD
system can become chaotic (as in
e.g. Ref.~\citenum{kuramoto1978diffusion}), or if this irregularity is
the result of a stochastic but nonchaotic process.
%

@@

%% biology~\cite{lawson2013spatial, robb2014stochastic,
%%   oguz2014stochastic, fu2013deterministic, rudiger2014stochastic}, and
In this paper we use Huang's original 2003
model~\cite{huang2003dynamic}, which has been widely
used~\cite{hattne2005stochastic,huang2004min,kerr2006division} in both
deterministic and stochastic variants, to study flattened cells of
$0.4\micron$ thickness that are similar to those observed by M\"annik
\emph{et al.}~\cite{mannik2009bacterial}.

@@

We mean to investigate the geometric limits of the Min system
oscillations as observed by M\"annik \emph{et
  al.}~\cite{mannik2012robustness}, so we have modeled the Min system
in several cell shapes and sizes.  Here we present a selection of
these, beginning with naturally occurring pill-shaped cells, followed
by a number of flattened out shapes which reflect the experiments of
M\"annik \emph{et al.}, in which bacteria are confined within thin
slits.  These slits are fabricated with a width of .25\micron, but
they are coated with a PDMS lining that the cells are able to deform,
raising uncertainty in the actual cell thickness.  Tests with pure
silicon, non-deformable slits show that the smallest thickness for
which cells are able to penetrate is
.4\micron~\cite{mannik2012robustness,mannik2009bacterial}.  We
therefore assume that the PDMS slits have been deformed to this
thickness and simulate flattened cells with a thickness of .4\micron.
Viewed from the top down the cells have the shapes described below and
viewed from the side they have at their edges a semicircular
cross-section (one may imagine the shape of a pancake).
%
In this paper we focus on four specific flattened cell shapes.  Two of
these shapes replicate those published by M\"annik \emph{et al.}, and
the other two are `stadium' shapes that respectively have the same
aspect ratio, thickness, and volume as the two cell shapes
experimentally observed by M\"annik \emph{et
  al.}~\cite{mannik2012robustness}.  Viewed from the top down, these
stadium shapes appear as rectangles with semi-circular end caps on the
long axis ends.

@@

Figure~\ref{image-p} shows a series of color plots of the density of
proteins at each stage of the reaction cycle. Deterministic simulation
data is shown above and stochastic simulation data is shown below.
The cells shown are $4\micron$ in length, measured from end to end.
We have `smeared out' the stochastic concentrations in a manner meant
to reproduce the images shown by diffraction limited fluorescence
microscopy.  We do so using the two dimensional Gaussian approximation
developed by Zhang \emph{et al.}~\cite{zhang2007gaussian}.  In this
approximation we use a numerical aperture value of 1.3, which is the
same as used by M\"annik experimentally, and a wavelength of $650nm$.
Each frame is 2.5 seconds ahead of the last, and each image shows the
concentration of a given state of protein (of the five described in
the reaction model) summed over the coordinate normal to the page.

@@

\begin{figure*}
  \centering
  \hspace{-3cm}
  \vspace{1cm}
  \includegraphics[width=16cm]{../paper/plot-ave}
  \caption{We display here arrows depicting successive maxima in space
    and and time overlayed on a color plot of the total MinD density
    averaged over the same time period.  The simulation time covered
    for the flattened cells is 350 seconds, which is the same
    period of time depicted in the experimental data plots of M\"annik
    \emph{et al.}~\cite{mannik2012robustness}.  For the wild-type pill
    shape, we only cover 250 seconds, in order to provide a
    useful comparison due to its shorter oscillation period.  The top
    row shows plots published by M\"annik \emph{et al.} of the MinD maxima behavior
    and the bottom two rows show our simulations using the stochastic
    and deterministic models, respectively.  We simulated
    approximations to the two shapes observed by M\"annik, which we call
    \emph{shape A} and \emph{shape B}.  In addition, we studied two
    flattened stadium shapes which we call \emph{stadium A} and
    \emph{stadium B} corresponding in aspect ratio and thickness to
    the two experimental shapes.  The spatial length scale of all the
    figures shown was identical. Each of the flattened cell shapes
    uses the same color scale for the number of proteins per unit
    area.  Finally, we display the natural pill shape, which was also
    featured in Figs.~\ref{image-p} and~\ref{corr-pill}, with a
    different color scale to reflect the thicker cell containing more
    proteins per cross-sectional area.  }
  \label{randst-plot-ave}
\end{figure*}

@@

\subsection*{Comparison of Experimental and Stadium Shapes}
%\label{sec:model-method-shapes}
As explained above, we focus on simulation of four illustrative
flattened cell shapes: two shapes created to replicate those shapes
observed experimentally by M\"annik \emph{et
  al.}~\cite{mannik2012robustness} (shape~\emph{A} and
shape~\emph{B}), and for comparison two `stadium' shapes that have the
identical aspect ratio, thickness and volume (stadium \emph{A} and
stadium \emph{B}), shown in Figure~\ref{randst-plot-ave}.  All of the
flattened cells have lost the rotational symmetry of the original pill
shapes, but while shape~\emph{A} and shape~\emph{B} have only one
mirror plane symmetry (that is normal to the flattened plane),
stadium~\emph{A} and stadium~\emph{B} have three.  This allows us to
distinguish between the effect of flattening the cell and the cell
shape's irregularity and asymmetry.

@@

From each of these flattened simulations, we have chosen a typical 350
second segment to compare with the published results of
M\"annik \emph{et al.}~\cite{mannik2012robustness}, exemplified in arrow plots in
which the arrow heads show the location of sequential MinD maxima
within the cell (in Fig.~\ref{randst-plot-ave}).  In addition to
arrows between successive maxima in space and time, we plot as a
colored background the density of MinD proteins averaged over the same
time period.  We note that we have manually verified that our
(computer-generated) arrow plots also reflect a human interpretation
of a movie of the same data.  Finally, for comparison we present the
same plot for a wild-type cell, with a time period of 250 seconds to
account for its short period of oscillation.

@@

%% We found that the deterministic simulations of the MinD system results
%% in robust and regular oscillatory behavoir of polar selection and
%% oscillation in not only symmetrical shapes, such as the wildtype pill
%% shape, but also in very assymetrical, flatttened cells, such as those
%% observed by M\"annik.  We also examined larger and smaller versions of
%% these shapes, and found this behavior to be very robust, from the
%% minimum size to sustain oscillation (around \fixme{???} $\mu$m) up to
%% around twice the cell size reported by M\"annik.  At larger sizes than
%% this, less regular oscillations occur using the deterministic method,
%% but we note that the cells reported by M\"annik are already considerably
%% larger in volume than typical wild type
%% cells~\cite{kubitschek1990cell,kubitschek1968linear,mannik2012robustness}.

@@

We find that flattened but regular and symmetric cells exhibit MinD
oscillations that are qualitatively similar to the oscillations
observed experimentally (and in our simulation) in irregular and
asymmetric cells.  This demonstrates that asymmetry is not required in
order to induce spatially irregular MinD oscillations.  In addition,
the temporal regularity of oscillations is only moderately affected by
irregularity in the shape of flattened cells.  Thus we conclude that
it is the flattening of the cells, and the accompanying expansion in the
transverse direction, rather than their irregular shape that primarily
causes the disruption of regular bipolar oscillation which is observed
by M\"annik \emph{et al.}~\cite{mannik2012robustness}.
